---
title: "Coursera Statistical Inference Assignment"
author: "Paolo Coraggio"
date: "28/10/2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=2, warning=FALSE, message=FALSE)
```

# Part II - Analysing the ToothGrowth dataset

The ToothGrowth dataset contains the data originally used to check the "Effect of Vitamin C on Tooth Growth in Guinea Pigs". The data measures the "response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice or ascorbic acid (a form of vitamin C and coded as VC)" (for futher details you can use `?ToothGrowth` R command).

## Visual inspection of the data and first summary

`ToothGrowth` dataframe has the following structure:

```{r ToothGrowth, echo=FALSE}
# Information about the dataframe
str(ToothGrowth)
unique(ToothGrowth$dose)
```

`str` function gives some basic information about the data.frame structure. We have three variables (len, supp, dose): `len` represent the tooth length, `supp` is a 2 level factor variables with "OJ" and "VC" the different method of vitamin C administration ("OJ" stands for Orange Juice and "VC" for ascorbic acid) and finally `dose` is the numeric value for dosage (possibly, 0.5, 1.0 or 2.0 mg/day).


## Statistical summary and visual inspection

Table 4. show the statistical summary of the ToothGrowth data frame about the recorded Tooth length by supplement and dose

```{r, echo=FALSE}
library(dplyr)
a <- ToothGrowth %>%
        group_by(supp, dose) %>%
        summarise(mean = mean(len), std = round(sd(len),2))

#knitr::kable(ptable, caption = "Exponential Vs Uniform - different sample size", col.names = c("Distribution",                                                              "N Samples",                                             "Mean",                                                              "Computed Sd",             "Expected Sd"))


knitr::kable(a, 
             caption = "Mean and Standard Deviation of Tooth length for dose and supplement",
             col.names = c("supplement", "dose (mg/day)", "mean", "sd"))
```

And the following plot shows these data.

```{r, fig.height=2.8, fig.align='center', echo=FALSE}
library(ggplot2)
library(ggpubr)

g3 <- ggplot(ToothGrowth, aes(x = dose, y = len, color = factor(dose))) +
    geom_boxplot(aes(factor(dose), len), alpha = 0.8) +
    theme(legend.position="right") + 
    scale_colour_discrete(name="dose")


g1 <- ggplot(ToothGrowth, aes(supp, len, color = supp)) +
    geom_boxplot(aes(factor(supp), len)) +
    theme(legend.position="right") +
    facet_wrap(.~dose) 
    
g2 <- ggplot(ToothGrowth, aes(dose, len, color = supp)) +
    geom_boxplot(aes(factor(dose), len)) +
    theme(legend.position="right") +
    facet_wrap(.~supp)

figure <- ggarrange(g3, g1, labels = c("A", "B"))
annotate_figure(figure,
                top = text_grob("Tooth growth by dose and supplement"),
                bottom = text_grob("Figure 2: boxplot of A) tooth growth by dose B) by supplement", hjust = 1, x = 1, size = 9))

```


Visually, we can guess that 

1. The tooth growth increase by administrating vitamin C
2. The supplement dosage has influence on the tooth growth
3. The way of administrating the supplement (OJ vs VC) seems to differ only when the dose is 1 mg/day otherwise the effect is similar

To confirm or reject these first clues based of visual comparison, I will perform statistical tests on the available data. 

## Tooth growth and dose

Plot A of figure 2. shows the tooth growth related to different dosage of vitamin C regardless of the supplement used. Since there is no recording of data of tooth growth without giving any vitamin C to Guinea Pigs (control), the $0.5 mg/day$ dosage will serve as baseline for comparison with other dosage.

The parameters for the t-test are the following:

* compare the recorded tooth lenght for each different dosage
* two sided t-test is used as the central values are compared (Figure B1 in Appendix B shows the histogram of the different datasets)
* the datasets are considered not paired since the subject receiving the dosages are different
* the variance of the dataset are considered not equal since we can't make any detailed assumption on the dataset

These are the result of the different t-test

```{r, echo = FALSE}
#head(ToothGrowth)

doses <-unique(ToothGrowth$dose)
combination <- combn(doses,2)

p.vals.dose <- c()
dose.conf.low <- c()
dose.conf.high <- c()

for(i in 1:dim(combination)[2])
{
    p.vals.dose <- c(t.test(ToothGrowth$len[ToothGrowth$dose == combination[1,i]],
           ToothGrowth$len[ToothGrowth$dose == combination[2,i]], 
           paired = FALSE, var.equal = FALSE)$p.value, p.vals.dose)
    
    dose.conf.low <- c(dose.conf.low, 
                       t.test(ToothGrowth$len[ToothGrowth$dose == 
                                                  combination[1,i]],
                              ToothGrowth$len[ToothGrowth$dose ==
                                                  combination[2,i]], 
                              paired = FALSE, var.equal = FALSE)$conf.int[1])
    
    dose.conf.high <- c(dose.conf.high, 
                       t.test(ToothGrowth$len[ToothGrowth$dose ==
                                                  combination[1,i]],
                              ToothGrowth$len[ToothGrowth$dose ==
                                                  combination[2,i]], 
                              paired = FALSE, var.equal = FALSE)$conf.int[2])
}

df_dose <- data.frame(dose.1 = combination[1,],
                dose.2 = combination[2,],
                p.values = round(p.vals.dose, 8),
                conf.int.low = round(dose.conf.low,2),
                conf.int.high = round(dose.conf.high,2))

# knitr::kable(df_dose)

knitr::kable(df_dose, caption = "Tooth Growth by Dose", col.names = c("dose 1",
                                                                      "dose 2",
                                                              "p value",
                                                              "left confidence interval",
                                                              "right confidence interval"), )

```

## Tooth growth and supplement

With reference to plot C, a t-test comparing the different supplement for each different dose is performed (i.e. comparing supplement "OJ" vs "VC" for the $0.5 mg/day$,  $1.0 mg/day$,  $2.0 mg/day$ doses respectively). The parameters of the 2 sided t-test are similar to the previous case.


```{r, echo=FALSE}
library(gridExtra)

p.vals.supp <- c()
supp.conf.low <- c()
supp.conf.high <- c()

for(i in doses)
{
    p.vals.supp <- c(p.vals.supp,
                     t.test(ToothGrowth$len[ToothGrowth$supp == "VC" &
                                                ToothGrowth$dose == i],
                            ToothGrowth$len[ToothGrowth$supp == "OJ" &
                                                ToothGrowth$dose == i],
                            paired = FALSE, var.equal = FALSE)$p.value)

    supp.conf.low <- c(supp.conf.low,
                     t.test(ToothGrowth$len[ToothGrowth$supp == "VC" &
                                                ToothGrowth$dose == i],
                            ToothGrowth$len[ToothGrowth$supp == "OJ" &
                                                ToothGrowth$dose == i],
                            paired = FALSE, var.equal = FALSE)$conf.int[1])

    supp.conf.high <- c(supp.conf.high,
                       t.test(ToothGrowth$len[ToothGrowth$supp == "VC" &
                                                  ToothGrowth$dose == i],
                              ToothGrowth$len[ToothGrowth$supp == "OJ" &
                                                  ToothGrowth$dose == i],
                              paired = FALSE, var.equal = FALSE)$conf.int[2])
}

df_supp <- data.frame(dose = c(0.5,1.0,2.0),
                      p.values = round(p.vals.supp,3),
                      conf.int.low = round(supp.conf.low,2),
                      conf.int.high = round(supp.conf.high,2))

knitr::kable(df_supp, caption = "Tooth Growth by Supplement", col.names = c("dose",
                                                              "p value",
                                                              "left confidence interval",
                                                              "right confidence interval"), )
```

## Conclusions

Table 5. shows that the mean of recorded length for the different dosage does not lie in the confidence intervals showing a very low p-value. The comparison between $0.5 mg/day$ versus $2.0  mg/day$ shows the major differences in mean and lowest p-value bringing to the conclusion that Vitamin C administration influences Tooth Growth.

Table 6. shows low p-values for low dosages while for $2.0$ dose, the supplement looks irrelevant ot Tooth grow.


# Appendix B - Part II, ToothGrowth

## Code

The full code is available on GitHub at the following [link](https://github.com/pacoraggio/CourseraAssignmentCode): https://github.com/pacoraggio/CourseraAssignmentCode 

## Plots

```{r, fig.align="centre", fig.height=3.5, echo=FALSE}
g <- ggplot(ToothGrowth, aes(len, color = supp)) +
    geom_histogram(binwidth = 2.5, color = "black", aes(fill = supp)) +
    facet_wrap(supp~dose)

g + labs(title = "Data disitribution", caption = "Figure B1: data distribution of length tooth growth")
```

